---
title: "PSTAT 174 Homework 5"
author: "Lucas Pieper-White"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1: Glossary Creation

# House Keeping 

```{r}
# set the working directory
setwd("path/to/folder") 

# print the current working directory
getwd()  
```
                        

# Reading Data

```{r}
# Read plain text from a file or URL
births <- scan("http://robjhyndman.com/tsdldata/data/nybirths.dat")

# Read CSV/tabular data: sep = delimiter, skip = rows to skip, nrows = max rows to read
wine.csv <- read.table("file.csv", sep=",", header=FALSE, skip=1, nrows=187)

# Read Excel files (requires readxl package)
library(readxl)
df <- read_excel("file.xlsx", skip=1, n_max=391)

# Preview a data frame
head(df)

# Get sample size
length(births)

```

# Creating Time Series Objects

```{r}
# Convert numeric vector to ts object: start = c(year, period), frequency = observations per year
#     frequency=4 for quarterly, =1 for annual
births_ts <- ts(births, start=c(1946,1), frequency=12)   # monthly

```


# Plotting Time Series

```{r}
# Plot ts object vs time axis
plot(births_ts, main="Time Series ____________")

# Alternative time series plot
ts.plot(births_ts)

# Plot data vs integer index 1, 2, ..., n
plot(1:length(births), births, type='l',
     xlab='index',
     main="Time Series ____________")

```


# Adding Tend and Mean lines

```{r}
# Create integer index vector
index <- 1:length(births)

# Fit OLS linear trend
trend <- lm(births ~ index)

# Overlay fitted trend line (red)
abline(trend, col="red")

# Overlay horizontal mean line (blue)
abline(h=mean(births), col="blue")

# Add dashed reference line at zero
abline(h=0, lty=2)

```


# Yearly/ Quarterly Aggregation:

```{r}
# Aggregate to yearly averages
# Also available: apply.monthly(), apply.quarterly(), apply.weekly(), apply.daily(). FUN can be mean, sum, or var
library(xts)
ts.yearly <- apply.yearly(as.xts(births_ts), FUN=mean)

# Plot aggregated data
plot(ts.yearly, main="Yearly Aggregated Time Series")

```


# Smoothing and Moving Averages

```{r}
library(zoo)
# 5-period rolling moving average
smooth_5m <- zoo::rollmean(births_ts, k=5, fill=NA)

# 12-period rolling moving average
smooth_12m <- zoo::rollmean(births_ts, k=12, fill=NA)

# Overlay on original plot
plot(births_ts, main="Original with Moving Averages")
lines(smooth_5m, col='blue')
lines(smooth_12m, col='red')
legend("topleft", legend=c("Original","5-month avg","12-month avg"),
       col=c("black","blue","red"), lty=1, cex=0.8)

# Symmetric convolution filter: X_t = (Z_{t-1} + Z_t + Z_{t+1}) / 3
x_t <- filter(z_t, filter=rep(1/3,3), sides=2, method="convolution")

```

# Decomposition Commands

```{r}
# Additive decomposition into trend, seasonal, and random
components <- decompose(births_ts)

# Plot all four decomposition panels
plot(components)

```

# White Noise Simulation and RV's

```{r}
# Generate n Gaussian white noise values ~ N(0,1)
wn <- rnorm(1000, mean=0, sd=1)

# Plot white noise
plot(1:length(wn), wn, type='l', main="White Noise", xlab='t', ylab='Z_t')

# Generate n uniform random variables on (min, max)
u <- runif(1000, min=-1, max=1)
mean(u)

# Simulate random walk with drift delta
z_t <- rnorm(499, 0, 1)
x_t <- c(0, cumsum(z_t))           # cum sum of WN
delta <- 0.6
rw <- 0:499 * delta + x_t          # X_t=delta*t + X_0 + sum(Z_j)
plot(rw, type="l", xlab="t", ylab="X_t", main="Random Walk with Drift")

```


# ARMA Simulation

```{r}
# Simulate MA(2): X_t = Z_t + 0.3*Z_{t-1} + 0.5*Z_{t-2}
ma_sim <- arima.sim(n=1000, model=list(ma=c(0.3, 0.5)))

# Simulate AR(1): X_t = 0.5*X_{t-1} + Z_t
ar_sim <- arima.sim(n=1000, model=list(ar=c(0.5)))

# Simulate AR(2): X_t = 0.7*X_{t-1} - 0.1*X_{t-2} + Z_t
ar2_sim <- arima.sim(n=200, model=list(ar=c(0.7, -0.1)))

# Simulate ARMA(1,1): X_t = 0.5*X_{t-1} + Z_t + 0.4*Z_{t-1}
arma_sim <- arima.sim(n=200, model=list(ar=c(0.5), ma=c(0.4)))

```

# SAMPLE ACF and PACF:

```{r}
# Compute and plot sample autocorrelation function
acf(x, lag.max=60, main="Sample ACF")

# Compute and plot sample partial autocorrelation function
pacf(x, lag.max=60, main="Sample PACF")

# Handle NAs in ACF computation
acf(x_t, na.action=na.pass, main="ACF with NAs")

```


# ACF and PACF (THEORETICAL)

```{r}
# Theoretical ACF for an AR model
theo_acf <- ARMAacf(ar=c(0.7,-0.1), lag.max=20, pacf=FALSE)

# Theoretical PACF for an AR model
theo_pacf <- ARMAacf(ar=c(0.7,-0.1), lag.max=20, pacf=TRUE)

# Theoretical ACF for an MA model
theo_acf_ma <- ARMAacf(ma=c(0.7,-0.1), lag.max=20, pacf=FALSE)

# Theoretical ACF for a mixed ARMA model
theo_acf_arma <- ARMAacf(ar=0.5, ma=0.4, lag.max=20, pacf=FALSE)

# Plot theoretical ACF as vertical bars
plot(theo_acf, type="h", ylim=c(-1,1), main="Theoretical ACF",
     col="red", ylab=expression(rho[X](h)), xlab="h")
abline(h=0)

# Overlay theoretical ACF (red) on top of sample ACF plot
acf(ar2_sim, lag.max=20, main="Sample ACF")
lines(x=0:20, y=theo_acf, col="red")
```

# Polynomial Root Commands for Causality and Invertibility

```{r}
# AR polynomial: phi(z) = 1 - 0.7z + 0.1z^2
# All |roots| > 1 => causal
polyroot(c(1, -0.7, 0.1))

# MA polynomial: theta(z) = 1 + 0.4z
# All |roots| > 1 => invertible
polyroot(c(1, 0.4))

# Compute absolute value of complex roots to check |z| > 1
roots <- polyroot(c(1, -0.75, 0.5625))
Mod(roots)

# Plot roots on the complex plane with unit circle
# (requires sourcing plot.roots.R from course materials)
source("plot.roots.R")
plot.roots(NULL, polyroot(c(1, -0.4004, 0.2866)), main="Roots of SAR part")
```

# Box Cox Transformations

```{r}
library(MASS)
t <- 1:length(wine)
fit <- lm(wine ~ t)
# Compute Box-Cox log-likelihood and plot to find optimal lambda
bcTransform <- boxcox(wine ~ t, plotit=TRUE)

# Extract optimal lambda value
lambda <- bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
lambda

# Apply Box-Cox transformation (use log(x) if lambda is approximately 0)
wine.bc <- (1/lambda) * (wine^lambda - 1)

# Compare original vs transformed
par(mfrow=c(1,2))
ts.plot(wine, main="Original", ylab=expression(X[t]))
ts.plot(wine.bc, main="Box-Cox Transformed", ylab=expression(Y[t]))
par(mfrow=c(1,1))

# Check normality of transformed data via histogram
par(mfrow=c(1,2))
hist(wine, main="Original")
hist(wine.bc, main="Box-Cox Transformed")
par(mfrow=c(1,1))

```

# Differencing Commands

```{r}
# First difference: (1-B)X_t, removes trend
y1 <- diff(x, 1)

# Seasonal difference at lag 12: (1-B^12)X_t, removes monthly seasonality
y_12 <- diff(x, 12)

# Both: first seasonal differencing, then first differencing
y1 <- diff(wine.bc, 12)
y12 <- diff(y1, 1)

# Track sample variance at each step — stop when variance starts increasing
var(wine.bc)  # after transformation
var(y1)       # after seasonal differencing
var(y12)      # after additional first differencing

```

# ARMA Fitting - MLE

```{r}
# Fit ARIMA(p,d,q) via maximum likelihood
fit <- arima(x, order=c(2,1,0), method="ML")
fit

# Extract AIC
AIC(fit)

# Extract coefficients and residuals
fit$coef
fit$residuals
```

# SARIMA Fitting

```{r}
library(astsa)
# Fit SARIMA(p,d,q) x (P,D,Q)_s
fit.s <- sarima(xdata=prodn, p=2, d=1, q=0,
                P=2, D=1, Q=2, S=12, details=FALSE)

# View full fitted arima object
fit.s$fit

# Extract corrected AIC (preferred for model comparison)
fit.s$AICc

# Extract coefficients
fit.s$fit$coef

# Plot residuals
plot(fit.s$fit$residuals)

# Fix insignificant coefficients to 0 (NA = estimate freely, 0 = fix at zero)
fit.s2 <- sarima(xdata=prodn, p=2, d=1, q=0,
                 P=2, D=1, Q=2, S=12, details=FALSE,
                 fixed=c(NA, 0, NA, NA, NA, NA))

```

# Yule Walker Estimation

```{r}
library(astsa)
# Fit SARIMA(p,d,q) x (P,D,Q)_s
fit.s <- sarima(xdata=prodn, p=2, d=1, q=0,
                P=2, D=1, Q=2, S=12, details=FALSE)

# View full fitted arima object
fit.s$fit

# Extract corrected AIC (preferred for model comparison)
fit.s$AICc

# Extract coefficients
fit.s$fit$coef

# Plot residuals
plot(fit.s$fit$residuals)

# Fix insignificant coefficients to 0 (NA = estimate freely, 0 = fix at zero)
fit.s2 <- sarima(xdata=prodn, p=2, d=1, q=0,
                 P=2, D=1, Q=2, S=12, details=FALSE,
                 fixed=c(NA, 0, NA, NA, NA, NA))

```

# Splitting Data- Train and Test

```{r}
# Remove first 24 observations (e.g., different pattern)
x_split <- births[-c(1:24)]

# Use first 132 obs for training
training <- x_split[1:132]

# Reserve last 12 obs for testing
testing <- x_split[133:144]

```

# General Plotting 
```{r}
# Set up side-by-side plots
par(mfrow=c(1,2))

# Reset to single plot
par(mfrow=c(1,1))

# Overlay a line on the current plot
lines(x, col="blue")

# Add legend
legend("topright", c("A","B"), col=c("black","blue"), lty=1, cex=0.8)

# Math expressions in labels
ylab = expression(X[t])                                    # X_t
ylab = expression(nabla^{12}~Y[t])                         # nabla^12 Y_t
main = expression(theta[1] == 0.45~theta[2] == 0.55)       # theta_1 = 0.45

```

